A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-05 podc_paper/algorithm.tex
Finished draft of analysis section.  (For some reason, compiling on my laptop requires commenting out some \newtheorem commands in podc-submission.tex .)
working on description of GC
updates based on Hossein's comments of Jan 4

To propagate an operation to a node $v$ in the tree, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a CAS instruction.
Following \cite{}, we call this a 3-step sequence a
\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A successful \op{Refresh} by $P$ may propagate pending operations by several other processes to $v$.
This cooperative approach is sufficient to ensure that if $P$ fails to CAS a 
new block into $v$'s array twice,
then its operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the root.

@@

The primary challenge is thus figuring out what information to store in each block so that 
the following tasks can be done efficiently (in a polylogarithmic number of steps).
\begin{enumerate}[label={(T\arabic*)}]
\item
\label{construct}
Construct a block for node $v$ that represents the operations contained in $O(p)$ consecutive blocks in $v$'s children, as required for a \op{Refresh}.
\item
\label{findinroot}
Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
\item
\label{findrank}
Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
or determine the rank $r$ of the enqueue whose value it should return.
\item
\label{findenqueue}
Find the $r$th enqueue in the linearization ordering.
\end{enumerate}
Since these tasks depends on the linearization ordering, we describe that ordering next.

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still in progress.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

@@

We shall show that an operation terminates only after a block containing it has been appended to the root's blocks array.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, our linearization orders operations according to the block they belong to in the root's blocks array.
Operations that appear in the same block are necessarily concurrent, so we can choose how to order them.

Each block in a leaf represents a single operation.
Each block $B$ in an internal node $v$ results from merging
several consecutive blocks from each of $v$'s children.
The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
A block $B'$ in a descendant of $v$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
or a subblock of a direct subblock of $B$.
A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.

The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
so there is at most one operation per process.
Hence, a block represents at most $p$ operations in total.  
Moreover, we never append empty blocks, so 
each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.

As mentioned above, we are free to order operations within a block however we like.
For convenience, we order the enqueues and dequeues separately, and put the 
operations that were propagated from the left child before the operations from the right child.
More formally, we inductively define an sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from the left child of $v$
and $B^R_1,\ldots,B^R_r$ from the right child of $v$, then $B$'s operation sequences are defined by the concatenations 
\begin{eqnarray}
E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{eqnarray}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}
\newcommand{\hangbox}[1]{\parbox[t]{\dimexpr\linewidth-\algorithmicindent\relax}{%
    \setlength{\hangindent}{\algorithmicindent}%
    #1}}
\newcommand{\spac}{\medskip}
\newcommand{\assign}{\mbox{:=}}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} \fld{\{left, right\}}} \label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
    \EndIf
    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
    \State \var{\Call{Advance}{\var{v, h}}}\label{advance}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
It uses the double refresh idea described
above and invokes two \op{Refresh}es on \var{v} in Lines
\ref{firstRefresh} and \ref{secondRefresh}. 
If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
that propagated blocks from \var{v}'s children to \var{v}.
Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.

%\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
in \var{v}'s
children and append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v},
so \op{Refresh} terminates and returns true at line \ref{addOP}.
Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a new block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
from \var{h} to $\var{h}+1$
and to fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
other processes help keep \fld{head} up to date by also calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.

>>>>

\Eric{Reworded following parag after Hossein's comments of Jan 4}
To propagate operations from a node $v$'s children to $v$, $P$ first observes
the blocks in both of $v$'s children that are not already in $v$,
creates a new block by combining information from those blocks, and attempts to append this 
new block to $v$'s blocks array using a \op{CAS} instruction.
Following \cite{}, we call this a 3-step sequence a
\op{Refresh} on $v$. %(see Figure \ref{fig::propagstep}).
A \op{Refresh}'s \op{CAS} may fail if there is another concurrent \op{Refresh} on $v$.
However, since a successful \op{Refresh} propagates multiple pending operations 
from $v$'s children to $v$,
we can prove that if two \op{Refresh}es by $P$ on $v$ fail,
then $P$'s operation has been propagated to $v$ by some other process, so $P$ can continue 
onwards towards the~root.

@@

The primary challenge is thus figuring out what information to store in each block so that 
the following tasks can be done efficiently (in a polylogarithmic number of steps).
\begin{enumerate}[label={(T\arabic*)}]
\item
\label{construct}
Construct a block for node $v$ that represents the operations contained in consecutive blocks in $v$'s children, as required for a \op{Refresh}.
\item
\label{findinroot}
Given a dequeue in a leaf that has been propagated to the root, find that operation's position in the root's blocks array.
\item
\label{findrank}
Given a dequeue's position in the root, determine whether it is a null dequeue (i.e., whether the queue is empty when it is linearized)
or determine the rank $r$ of the enqueue whose value it should return.
\item
\label{findenqueue}
Find the $r$th enqueue in the linearization ordering.
\end{enumerate}
Since these tasks depends on the linearization ordering, we describe that ordering next.

\begin{figure}[tb]
\input{tournamentTree.pdf_t}
\caption{An example tournament tree with four processes. 
We show explicitly the enqueue sequence and dequeue sequence represented by each block in the \fld{blocks} arrays of the seven nodes.  The leftmost element of each \fld{blocks} array is a dummy block.
Arrows represent the indices stored in \eleft\ and \eright\ fields of blocks (as described in Section \ref{sec:fields}).
The fourth process's Deq\sub{6} is still propagating towards the root.
The linearization order for this tree is
Enq(a) Enq(e) Deq\sub{2} $\mid$ Enq(b) Deq\sub{4} Deq\sub{5} $\mid$ Enq(d) Enq(f) Enq(h) Deq\sub{1} $\mid$ Enq(c) Deq\sub{3} $\mid$ Enq(g), where vertical bars indicate boundaries of blocks in the root.\label{tournament}}
\end{figure}

@@

Performing a double refresh at each node along the path from the leaf to the root ensures 
a block containing the operation is appended to the root before the operation termintes.
So, if an operation $op_1$ terminates before another operation $op_2$ begins, 
$op_1$ will be in an earlier block than $op_2$ in root's blocks array.
Thus, we linearize operations according to the block they belong to in the root's blocks array.
We can choose how to order operations in the same block, since they must be concurrent.

Each block in a leaf represents one operation.
Each block $B$ in an internal node $v$ results from merging
several consecutive blocks from each of $v$'s children.
The merged blocks in $v$'s children are called the \emph{direct subblocks} of $B$.
A block $B'$ is a \emph{subblock} of $B$ if it is a direct subblock of $B$
or a subblock of a direct subblock of $B$.
A block $B$ represents the set of operations in all of $B$'s subblocks in leaves of the tree.
The operations propagated by a \op{Refresh} are all pending when the \op{Refresh} occurs,
so there is at most one operation per process.
Hence, a block represents at most $p$ operations in total.  
Moreover, we never append empty blocks, so 
each block represents at least one operation and it follows that a block can have at most $p$ direct subblocks.

As mentioned above, we are free to order operations within a block however we like.
We order the enqueues and dequeues separately, and put the 
operations propagated from the left child before the operations from the right child.
More formally, we inductively define sequences $E(B)$ and $D(B)$ of the enqueues and dequeues
represented by a block $B$.
If $B$ is a block in a leaf representing an enqueue operation, its enqueue sequence $E(B)$ is that operation
and its dequeue sequence $D(B)$ is empty.  If $B$ is a block in a leaf representing a dequeue, $D(B)$ is that single operation and $E(B)$ is empty.
If $B$ is a block in an internal node $v$ with direct subblocks $B^L_1, \ldots, B^L_\ell$ from 
$v$'s left child
and $B^R_1,\ldots,B^R_r$ from $v$'s right child, then $B$'s operation sequences are defined by the concatenations 
\begin{eqnarray}
E(B) &=& E(B^L_1)\cdots E(B^L_\ell)\cdot E(B^R_1) \cdots E(B^R_r)\nonumber\\
D(B) &=& D(B^L_1)\cdots D(B^L_\ell)\cdot D(B^R_1) \cdots D(B^R_r)\label{defSeqs}
\end{eqnarray}
We say the block $B$ \emph{contains} the operations in $E(B)$ and $D(B)$.

@@

\renewcommand{\algorithmiccomment}[1]{\hfill\eqparbox{COMMENTDOUBLE}{\com\ #1}}

\here{For consistency of notation, perhaps change $n$ to $B$ in code for Enqueue, Dequeue.  Maybe also new to B for CreateBlock, Refresh.}

@@

\Function{boolean}{Refresh}{\typ{Node} \var{v}}
    \State \linecomment{try to append a new block to \var{v}.\fld{blocks}}
    \State \var{h} \assign\ \var{v}.\head \label{readHead}
    \ForEach{\fld{dir} {\keywordfont{in}} $\{\fld{left, right}\}$} \label{startHelpChild1}
        \State \var{childHead} \assign\ \var{v}.\fld{dir}.\head \label{readChildHead}
        \If{\var{v}.\fld{dir.blocks}[\var{childHead}] $\neq$ \nl} \label{ifHeadnotNull}
            \State \Call{Advance}{\var{v}.\fld{dir}, \var{childHead}} \label{helpAdvance}
        \EndIf
    \EndFor \label{endHelpChild1}
    \State \var{new} \assign\ \Call{CreateBlock}{\var{v, h}} \label{invokeCreateBlock}
    \If{\var{new = \nl}} \Return{\tr} \label{addOP} 
    \EndIf
    \State \var{result} \assign\ \Call{CAS}{\var{v}.\fld{blocks}[\var{h}], \nl, \var{new}} \label{cas}
    \State \Call{Advance}{\var{v, h}}\label{advance}
    \State \Return{ \var{result}}
\EndFunction{Refresh}

@@

\opemph{Propagate}(\var{v}) guarantees that any blocks that are in \var{v}'s children when \op{Propagate} is invoked are propagated to the root.
It uses the double refresh idea described
above and invokes two \op{Refresh}es on \var{v} in Lines
\ref{firstRefresh} and \ref{secondRefresh}. 
If both fail to add a block to \var{v}, it means some other process has done a successful \op{Refresh}
that propagated blocks that were in \var{v}'s children prior to line \ref{firstRefresh} to \var{v}.
Then, \op{Propagate} recurses to \var{v}.\fld{parent} to continue propagating blocks up to the root.

%\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
in \var{v}'s
children and append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v}.
In this case, \op{Refresh} returns true at line \ref{addOP} and terminates.
Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a new block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
from \var{h} to $\var{h}+1$
and to fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
other processes help keep \fld{head} up to date by also calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.

### 2023-01-05 podc_paper/correctness.tex
working on description of GC
updates based on Hossein's comments of Jan 4

A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

>>>>

\Eric{I'm using immutable to mean that they are written only when the object is created.  The \fld{super} field does change from null to some value after the block is created, so it 
is written once, rather than 0 times.  In other words, we could have initialized
 all the other fields when the Block constructor is called.}
A \typ{Block} object's fields, except for \fld{super}, are immutable:  they are written only 
when the block is created at line \ref{enqNew} or \ref{deqNew} (for a leaf's block) or lines \ref{initNewBlock}--\ref{computeLength} (for an internal node's block).  
Moreover, only a \op{CAS} at line \ref{setSuper1} can modify the \fld{super} field 
(from \nl\ to a non-\nl\ value), so it remains fixed once a value is stored in it.
Similarly, only a \op{CAS} at line \ref{cas} can modify an element of a node's \fld{blocks} array 
(from \nl\ to a non-\nl\ value), so once a block is stored in a node, it remains there forever.
Only a \op{CAS} at line \ref{incrementHead} can update a node's \head\ field by incrementing it,
which implies the following.

@@

\here{Should there be a lemma somewhere in this section that explicitly shows that each
object we dereference is non-null?}
